Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-27 paper.tex
pics are smaller
caption for figure 1

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this behavior.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super)}
\end{figure}

>>>>

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback} (PLACEHOLDER DATA). The TDR threshold represents the duration after which the OS, assuming a non-terminating kernel, kills the driver. \john{This figure needs to be cited in the text.}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super).}
\end{figure}
